# R Nonlinear Regression

## Learning Objectives

In this chapter, you will learn how to:

- Run a regression with a quadratic term
- Run a regression with log transformations

## Set-up

To complete this chapter, you need to

- Start a R Markdown document
- Change the YAML to at least the following. Feel free to add arguments.

```{r, eval=FALSE}
---
title: 'R Chapter 7'
author: 'Your name'
output: 
  html_document:
    theme: spacelab
    df_print: paged
---
```
- Load the following packages

```{r, eval=FALSE}
library(tidyverse)
library(moderndive)
library(carData)
```

We will use the `Mroz` dataset within the `carData` package. Be sure to view the documentation for these data in the Help tab of the bottom-right pane by typing the name of the dataset in the search bar.

## Quadratic term

Recall the below regression model from Chapter \@ref(nonlinear-variables) that includes a squared term for `Age`, which allows our regression line to change directions once as `Age` changes. We included this term because Figure \@ref(fig:quadscatter) suggested wages initially increase with age, then decreases.

\begin{equation}
Wage = \beta_0 + \beta_1Age + \beta_2Age^2 + \beta_3Educ + \epsilon
\end{equation}

The below code demonstrates how to include a quadratic term within the `lm` function.

```{r}
quad_mod <- lm(Wage ~ Age + I(Age^2) + Educ, data = wages)
```

In this case, the code reflects the equation only somewhat; the `I()` is necessary to tell R that `Age^2` is the squared version of `Age`. Otherwise, R would not recognize `Age^2` in the data, thus excluding it from the regression.

Now we can obtain results in the usual manner.
```{r, eval=FALSE}
get_regression_table(quad_mod)
```

```{r, echo=FALSE}
get_regression_table(quad_mod) %>% 
  kable(format = 'html')
```

We need to alter the `Mroz` data slightly before running a regression. Run the following code that creates a new variable that equals 1 if `lfp` equals "yes" and 0 if `lfp` equals "no." This is necessary because our outcome variable--even though categorical--must be represented numerically in order for the regression to work.

```{r, eval=FALSE}
my_Mroz <- Mroz %>% 
  mutate(lfp_numeric = if_else(lfp == "yes", 1, 0))
```


```{block, type='learncheck', purl=FALSE}
**Exercise 1:** Suppose we want to examine factors that explain whether married women participate in the labor force, which is a binary outcome. We use the following model:

\begin{equation}
lfp = \beta_0 + \beta_1k5 + \beta_2age + \beta_3age^2 + \beta_4wc + \beta_5lwg + \beta_6inc + \epsilon
\end{equation}

Run this regression model and obtain the results.
```

## Log Transformation

In Chapter \@ref(nonlinear-variables), the following log-log regression model was run.

\begin{equation}
ln(LifeExp)=\beta_0 + \beta_1ln(GDPpercap) + \beta_2Continent + \epsilon
\end{equation}

The below code demonstrates how to transform a variable into its natural log within the `lm` function.

```{r}
loglog <- lm(log(lifeExp) ~ log(gdpPercap) + continent, data = gapminder)
```

Note that all we need to do is place the appropriate variables within the `log()` function, which R interprets as the natural log. This *temporarily* transforms the variables; it does not create new variables in the dataset equal to the natural log of the variables.

Now we can obtain results in the usual manner.
```{r, eval=FALSE}
get_regression_table(loglog)
```

```{r, echo=FALSE}
get_regression_table(loglog) %>% 
  kable(format = 'html')
```

```{block, type='learncheck', purl=FALSE}
**Exercise 2:** Suppose we decide we want to use the natural log of family income exclusive of wife's income, `inc`, resulting in the following model

\begin{equation}
lfp = \beta_0 + \beta_1k5 + \beta_2age + \beta_3age^2 + \beta_4wc + \beta_5lwg + \beta_6ln(inc) + \epsilon
\end{equation}

Run this regression model and obtain the results.
```

## Save and Upload

Knit your Rmd to save it and check for errors. If you are satisfied with your work, upload to eLC. Once you upload, answers will become available for download.